---
title: "Breast Cancer Wisconsin (Diagnostic) — Data Analysis & Decision Tree"
author: "Beatriz Martins Gomes Silva"
format:
  html:
    css: styles.css
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

## 1. Context & objective

This report presents an end-to-end workflow on the **Breast Cancer Wisconsin (Diagnostic)** dataset:

- Understand the dataset structure and quality
- Explore distributions and relationships
- Build a first **baseline** classification model (Decision Tree)
- Evaluate performance on a train/test split

> Target variable: `diagnosis` (`M` = malignant, `B` = benign)

## 2. Setup

```{r}
# Core
library(tidyverse)
library(janitor)

# Modeling
library(caret)
library(rsample)
library(rpart)
library(rpart.plot)
library(yardstick)
library(pROC)
```

```{r}
# --- Style (plots) ---
bronze <- "#CD7F32"
bronze_dark <- "#8C5A2B"
bronze_light <- "#F5E6D3"

theme_set(theme_classic(base_size = 14, base_family = "Times New Roman") +
            theme(
              plot.title = element_text(face = "bold"),
              plot.subtitle = element_text(size = rel(0.9)),
              plot.caption = element_text(size = rel(0.85), color = bronze_dark),
              legend.position = "top"
            ))
```


## 3. Load & clean data

```{r}
df_raw <- read_csv("data.csv", show_col_types = FALSE) |> clean_names()

glimpse(df_raw)
```

### 3.1 Basic cleaning

- Remove the empty column (`unnamed_32`)
- Ensure `diagnosis` is a factor with explicit labels

```{r}
df <- df_raw |>
  select(-any_of("unnamed_32")) |>
  mutate(
    diagnosis = factor(diagnosis, levels = c("B","M"), labels = c("Benign","Malignant"))
  )

skimr::skim(df)
```

> If `skimr` is not installed, you can remove the line above.

## 4. Exploratory Data Analysis (EDA)

### 4.1 Class balance

```{r}
df |>
  count(diagnosis) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = diagnosis, y = n, fill = diagnosis)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  scale_fill_manual(values = c("Benign" = bronze_light, "Malignant" = bronze_dark)) +
  labs(
    title = "Class balance",
    subtitle = "Benign vs malignant",
    x = NULL,
    y = "Number of samples",
    caption = "Class counts"
  )
```

### 4.2 Feature distributions (radius, perimeter, area)

```{r}
df_long <- df |>
  select(diagnosis, radius_mean, perimeter_mean, area_mean) |>
  pivot_longer(
    cols = c(radius_mean, perimeter_mean, area_mean),
    names_to = "feature",
    values_to = "value"
  )

df_long |>
  ggplot(aes(x = value, fill = diagnosis)) +
  geom_histogram(bins = 30, alpha = 0.75, position = "identity") +
  facet_wrap(~ feature, scales = "free", ncol = 1) +
  scale_fill_manual(values = c("Benign" = bronze_light, "Malignant" = bronze)) +
  labs(
    title = "Feature distributions",
    subtitle = "Radius, perimeter and area",
    x = NULL,
    y = "Count",
    caption = "Distributions by class"
  )
```

### 4.3 Correlation heatmap (readable subset)

```{r, fig.width=12, fig.height=10}
# The dataset contains many numeric features, which can make a full heatmap unreadable.
# To keep it interpretable, we show a subset (including the 3 key features used later).

vars_heatmap <- c(
  "radius_mean", "perimeter_mean", "area_mean",
  "texture_mean", "smoothness_mean", "compactness_mean",
  "concavity_mean", "concave_points_mean", "symmetry_mean",
  "fractal_dimension_mean", "radius_worst", "area_worst"
)

num_df <- df |> select(any_of(vars_heatmap))

corr <- cor(num_df, use = "pairwise.complete.obs")

corr_long <- as.data.frame(as.table(corr)) |>
  rename(var1 = Var1, var2 = Var2, corr = Freq)

corr_long |>
  ggplot(aes(x = var1, y = var2, fill = corr)) +
  geom_tile() +
  coord_equal() +
  scale_fill_gradient2(
    low = bronze_light,
    mid = "white",
    high = bronze_dark,
    midpoint = 0,
    limits = c(-1, 1),
    name = "r"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.text.y = element_text(size = 10)
  ) +
  labs(
    title = "Correlation heatmap",
    subtitle = "Pearson correlation (subset)",
    x = NULL,
    y = NULL,
    caption = "Scale: -1 to 1"
  )
```

## 5. Modeling — Decision Tree (3 features)

### 5.1 Train/test split

```{r}
set.seed(42)

split <- initial_split(df, prop = 0.7, strata = diagnosis)
train <- training(split)
test  <- testing(split)

table(train$diagnosis)
table(test$diagnosis)
```

### 5.2 Train model (radius, perimeter, area)

We train a decision tree using only **three features**:

- `radius_mean`
- `perimeter_mean`
- `area_mean`

```{r}
set.seed(42)

train_3 <- train |> select(diagnosis, radius_mean, perimeter_mean, area_mean)
test_3  <- test  |> select(diagnosis, radius_mean, perimeter_mean, area_mean)

ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

dt3 <- caret::train(
  diagnosis ~ .,
  data = train_3,
  method = "rpart2",
  tuneLength = 20,
  trControl = ctrl,
  metric = "ROC"
)

rpart.plot(
  dt3$finalModel,
  cex = 0.85,
  extra = 104,
  type = 1,
  box.palette = c(bronze_light, bronze_dark)
)
```

## 6. Evaluation

### 6.1 Predictions

```{r}
pred_class <- predict(dt3, newdata = test_3)
pred_prob  <- predict(dt3, newdata = test_3, type = "prob")[, "Malignant"]

results <- tibble(
  truth = test_3$diagnosis,
  pred  = pred_class,
  prob_malignant = pred_prob
)
```

### 6.2 Confusion matrix and metrics

```{r}
conf_mat(results, truth = truth, estimate = pred)

metrics(results, truth = truth, estimate = pred) |>
  filter(.metric %in% c("accuracy", "sens", "spec"))
```

### 6.3 ROC curve + AUC

```{r}
roc_obj <- pROC::roc(
  response = results$truth,
  predictor = results$prob_malignant,
  levels = c("Benign", "Malignant"),
  direction = "<"
)

plot(roc_obj, main = "ROC curve — Decision Tree (3 features)")
pROC::auc(roc_obj)
```

## 7. Rule-based classifier (manual tree)

To mirror the hand-made approach from the coursework, we implement a simple rule-based classifier:

- If `radius_mean > 15` → **Malignant**
- Else if `perimeter_mean > 100` → **Malignant**
- Else if `area_mean > 750` → **Malignant**
- Else → **Benign**

```{r}
rule_predict <- function(radius_mean, perimeter_mean, area_mean) {
  if (radius_mean > 15) {
    return("Malignant")
  }
  if (perimeter_mean > 100) {
    return("Malignant")
  }
  if (area_mean > 750) {
    return("Malignant")
  }
  return("Benign")
}

rule_pred <- pmap_chr(
  test_3 |> select(radius_mean, perimeter_mean, area_mean),
  rule_predict
)

rule_results <- tibble(
  truth = test_3$diagnosis,
  pred  = factor(rule_pred, levels = levels(test_3$diagnosis))
)

conf_mat(rule_results, truth = truth, estimate = pred)
metrics(rule_results, truth = truth, estimate = pred) |>
  filter(.metric %in% c("accuracy", "sens", "spec"))
```

### 7.1 Manual check on one sample

```{r}
one <- test_3[1, ]
one

rule_predict(one$radius_mean, one$perimeter_mean, one$area_mean)
```

## 8. Conclusions & next steps

This report combines an explainable decision tree model (3 features) with a simple rule-based classifier.

Potential next steps:
- Compare several models (logistic regression, random forest) as additional baselines
- Improve reporting of errors and misclassifications
- Add feature engineering and calibration when appropriate
